---
title: "What Factors Determine a Students Success?"
---


# What we did

## Chose Data:

-   we really wanted to find a data set that would apply to us and could give us results that we would care about

-   We were all interested to see how a students environment effected their success in school

-   Results could be used to paint a better picture for where external funding should go for students falling behind. 

## Clean Data

-   This data set was relatively small, only 1000 entries, so there really was not that much garbage to clean out

-   all exam scores ranged from 0-100, and every student had a full attribute list.


## Transform Data

```{r}
library(tidyverse) 

StudentPerformance <- readr::read_csv("./datas/StudentsPerformance.csv")

newSP <- StudentPerformance |>

  mutate(mean_score = (`math score` + `reading score` + `writing score`)/3) |>

  select(-`math score`, -`writing score`, -`reading score`) |>

  pivot_longer(cols = 1:5, names_to = "variable", values_to = "category")

```

-   Organize the data into a comprehensive structure

## Goals

-   The goal of our project was to use the data to determine which attributes of the students were most important

-   There are 5 attributes in this data set 

    -   Gender

    -   Ethnicity

    -   Education Level of the students parents

    -   Whether they were enrolled in some sort of school lunch program

    -   If they took any preparation courses for the exam

-   We needed to compare all of these to decide which one was the most important factor in each of the 3 subjects and overall. 

## Process Data

-   First we needed some baseline averages for each category so we have a reference to work with

-   We used R to go through each column and find an average score for each section, and a average total score

-   We then modeled each attribute to show its frequency in the data so we had a better idea as to what the population looked like

-   

```{r}
#| echo: true
#| eval: true

library(ggplot2)

students <- readr::read_csv("./datas/StudentsPerformance.csv")
readingMean <- mean(students$'reading score', na.rm = TRUE)
writingMean <- mean(students$'writing score', na.rm = TRUE)
mathMean <- mean(students$'math score', na.rm = TRUE)
```
```{r}
#| echo: false
print(students)
```

## Mean Scores

- Mean math score `r mathMean` 

- Mean Reading score `r readingMean`

- Mean writing score `r writingMean`

Now that we have all of our means, we can start seeing where the variations lie within attributes



# Effects of Prep Courses on scores
```{r}
library(ggplot2)



newSP |>

    filter(variable == 'test preparation course') |>

    ggplot(aes(y=mean_score, x = category, fill=category))+

    geom_boxplot(color = "black")+

    labs(y = "Average Score",x='', title = "Preparation Course")+

    theme_minimal()+

    theme(

    axis.title.x = element_blank(),

    axis.text.x = element_blank(),

    axis.ticks.x = element_blank()

    )
```

```{r}
#| echo: false

library(dplyr)
#| echo: false

MathByPrep <- students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

avg_math_completed <- MathByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_math_score)
avg_math_none <- MathByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_math_score)
```


```{r}
#| echo: false


library(dplyr)
ReadingByPrep<-students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

avg_reading_completed <- ReadingByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_reading_score)
avg_reading_none <- ReadingByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_reading_score)
WritingByPrep <- students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

avg_writing_completed <- WritingByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_writing_score)
avg_writing_none <- WritingByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_writing_score)

prep_levels <- c("Completed", "None")

avg_reading_scores <- c(avg_reading_completed, avg_reading_none)
avg_writing_scores <- c(avg_writing_completed, avg_writing_none)
avg_math_scores <- c(avg_math_completed, avg_math_none)

df <- tibble(
  Prep_Course = factor(rep(prep_levels, 3), levels = prep_levels), 
  Subject = rep(c("Reading", "Writing", "Math"), each = length(prep_levels)),  
  Score = c(avg_reading_scores, avg_writing_scores, avg_math_scores)  
)


```

## Results



-   There was a `r avg_reading_completed - avg_reading_none`% increase for reading scores when prep courses were taken

-   There was a `r avg_writing_completed - avg_writing_none`% increase for writing scores when prep courses were taken

-   There was a `r avg_math_completed - avg_math_none`% increase for math scores when prep courses were taken


# Effects of Gender on scores
```{r}
newSP |>

    filter(variable == 'gender') |>

    ggplot(aes(y=mean_score, x = category, fill = category))+

    geom_boxplot(color = "black")+

    labs(y = "Average Score",x='', title = "Gender", fill ='')+

    theme_minimal()+

    theme(

    axis.title.x = element_blank(),

    axis.text.x = element_blank(),

    axis.ticks.x = element_blank()

    )
```



```{r}
#| echo: false

ReadingByGender<-students %>%
  group_by(`gender`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByGender<-students %>%
  group_by(`gender`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByGender<-students %>%
    group_by(`gender`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

#print(MathByGender)
#print(ReadingByGender)
#print(WritingByGender)

avg_writing_male <- WritingByGender %>% filter(`gender` == "male") %>% pull(avg_writing_score)
avg_writing_female <- WritingByGender %>% filter(`gender` == "female") %>% pull(avg_writing_score)

avg_reading_male <- ReadingByGender %>% filter(`gender` == "male") %>% pull(avg_reading_score)
avg_reading_female <- ReadingByGender %>% filter(`gender` == "female") %>% pull(avg_reading_score)

avg_math_male <- MathByGender %>% filter(`gender` == "male") %>% pull(avg_math_score)
avg_math_female <- MathByGender %>% filter(`gender` == "female") %>% pull(avg_math_score)


genders <- c("Male", "Female")

avg_reading_scores <- c(avg_reading_male, avg_reading_female)
avg_writing_scores <- c(avg_writing_male, avg_writing_female)
avg_math_scores <- c(avg_math_male, avg_math_female)


df <- tibble(
  Gender = factor(rep(genders, 3), levels = genders),  # Set factor levels explicitly
  Subject = rep(c("Reading", "Writing", "Math"), each = length(genders)),  # Repeat subjects
  Score = c(avg_reading_scores, avg_writing_scores, avg_math_scores)  # Combine all scores
)

ggplot(df, aes(x = Gender, y = Score, fill = Subject)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Gender",
       x = "Gender",
       y = "Average Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Results

The effects of gender on test scores are very interesting

- In Mathematics, the males pull ahead by `r avg_math_male - avg_math_female`%

- In Reading and writing, the females win by a large gap

  - Reading: `r avg_reading_female - avg_reading_male`% win for females

  - Writing: `r avg_writing_female - avg_writing_male`% win for females

# Effects of ethnicity on scores

```{r}
#| echo: false

library(dplyr)

ReadingByEthnicity<-students %>%
  group_by(`race/ethnicity`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByEthnicity<-students %>%
  group_by(`race/ethnicity`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByEthnicity<-students %>%
    group_by(`race/ethnicity`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

avgMathEthnicityA<-MathByEthnicity %>% filter(`race/ethnicity` == "group A") %>% pull(avg_math_score)
avgMathEthnicityB<-MathByEthnicity %>% filter(`race/ethnicity` == "group B") %>% pull(avg_math_score)
avgMathEthnicityC<-MathByEthnicity %>% filter(`race/ethnicity` == "group C") %>% pull(avg_math_score)
avgMathEthnicityD<-MathByEthnicity %>% filter(`race/ethnicity` == "group D") %>% pull(avg_math_score)
avgMathEthnicityE<-MathByEthnicity %>% filter(`race/ethnicity` == "group E") %>% pull(avg_math_score)

avgReadingEthnicityA<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group A") %>% pull(avg_reading_score)
avgReadingEthnicityB<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group B") %>% pull(avg_reading_score)
avgReadingEthnicityC<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group C") %>% pull(avg_reading_score)
avgReadingEthnicityD<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group D") %>% pull(avg_reading_score)
avgReadingEthnicityE<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group E") %>% pull(avg_reading_score)

avgWritingEthnicityA<-WritingByEthnicity %>% filter(`race/ethnicity` == "group A") %>% pull(avg_writing_score)
avgWritingEthnicityB<-WritingByEthnicity %>% filter(`race/ethnicity` == "group B") %>% pull(avg_writing_score)
avgWritingEthnicityC<-WritingByEthnicity %>% filter(`race/ethnicity` == "group C") %>% pull(avg_writing_score)
avgWritingEthnicityD<-WritingByEthnicity %>% filter(`race/ethnicity` == "group D") %>% pull(avg_writing_score)
avgWritingEthnicityE<-WritingByEthnicity %>% filter(`race/ethnicity` == "group E") %>% pull(avg_writing_score)

EthAAvg = (avgMathEthnicityA + avgReadingEthnicityA + avgWritingEthnicityA)/3
EthBAvg = (avgMathEthnicityB + avgReadingEthnicityB + avgWritingEthnicityB)/3
EthCAvg = (avgMathEthnicityC + avgReadingEthnicityC + avgWritingEthnicityC)/3
EthDAvg = (avgMathEthnicityD + avgReadingEthnicityD + avgWritingEthnicityD)/3
EthEAvg = (avgMathEthnicityE + avgReadingEthnicityE + avgWritingEthnicityE)/3



ethnicities <- c("A", "B", "C", "D","E")
avg_reading_scores <- c(avgReadingEthnicityA, avgReadingEthnicityB, avgReadingEthnicityC, avgReadingEthnicityD, avgReadingEthnicityE)
avg_writing_scores <- c(avgWritingEthnicityA, avgWritingEthnicityB, avgWritingEthnicityC, avgWritingEthnicityD, avgWritingEthnicityE)
avg_math_scores <- c(avgMathEthnicityA, avgMathEthnicityB, avgMathEthnicityC, avgMathEthnicityD, avgMathEthnicityE)

df <- tibble(
  Ethnicity = factor(rep(ethnicities, 3), levels = ethnicities),  # Set factor levels explicitly
  Subject = rep(c("Reading", "Writing", "Math"), each = length(ethnicities)),  # Repeat subjects
  Score = c(avg_reading_scores, avg_writing_scores, avg_math_scores)  # Combine all scores
)

```

```{r}
newSP |>

    filter(variable == 'race/ethnicity') |>

    ggplot(aes(y=mean_score, x = category, fill=category))+

    geom_boxplot(color = "black")+

    labs(y = "Average Score",x='', title = "Race/Ethnicity")+

    theme_minimal()+

    theme(

    axis.title.x = element_blank(),

    axis.text.x = element_blank(),

    axis.ticks.x = element_blank()

    )
```

## Results

- Ethnicity A avg score : `r EthAAvg`

- Ethnicity B avg score : `r EthBAvg`

- Ethnicity C avg score : `r EthCAvg`

- Ethnicity D avg score : `r EthDAvg`

- Ethnicity E avg score : `r EthEAvg`


There is some variation between ethnicities on scores

Ethnicity E has a 9% advantage over Ethnicity A

# Effects of Parents Education Level on Scores
 there are in this data set: 
 some highschool, highschool, some college, associates, bachelors, masters



```{r}
#| echo: false

ReadingByEducation<-students %>%
  group_by(`parental level of education`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByEducation<-students %>%
  group_by(`parental level of education`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByEducation<-students %>%
    group_by(`parental level of education`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))



avgMathsomeHighschool<-MathByEducation %>% filter(`parental level of education` == "some high school") %>% pull(avg_math_score)
avgReadingsomeHighschool<-ReadingByEducation %>% filter(`parental level of education` == "some high school") %>% pull(avg_reading_score)
avgWritingsomeHighschool<-WritingByEducation %>% filter(`parental level of education` == "some high school") %>% pull(avg_writing_score)

avgMathsHighschool<-MathByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_math_score)
avgReadingHighschool<-ReadingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_reading_score)
avgWritingHighschool<-WritingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_writing_score)

avgMathHighschool<-MathByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_math_score)
avgReadingHighschool<-ReadingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_reading_score)
avgWritingHighschool<-WritingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_writing_score)

avgMathSomeCollege<-MathByEducation %>% filter(`parental level of education` == "some college") %>% pull(avg_math_score)
avgReadingSomeCollege<-ReadingByEducation %>% filter(`parental level of education` == "some college") %>% pull(avg_reading_score)
avgWritingSomeCollege<-WritingByEducation %>% filter(`parental level of education` == "some college") %>% pull(avg_writing_score)

avgMathAssociates<-MathByEducation %>% filter(`parental level of education` == "associate's degree") %>% pull(avg_math_score)
avgReadingAssociates<-ReadingByEducation %>% filter(`parental level of education` == "associate's degree") %>% pull(avg_reading_score)
avgWritingAssociates<-WritingByEducation %>% filter(`parental level of education` == "associate's degree") %>% pull(avg_writing_score)

avgMathBachelors<-MathByEducation %>% filter(`parental level of education` == "bachelor's degree") %>% pull(avg_math_score)
avgReadingBachelors<-ReadingByEducation %>% filter(`parental level of education` == "bachelor's degree") %>% pull(avg_reading_score)
avgWritingBachelors<-WritingByEducation %>% filter(`parental level of education` == "bachelor's degree") %>% pull(avg_writing_score)

avgMathMasters<-MathByEducation %>% filter(`parental level of education` == "master's degree") %>% pull(avg_math_score)
avgReadingMasters<-ReadingByEducation %>% filter(`parental level of education` == "master's degree") %>% pull(avg_reading_score)
avgWritingMasters<-WritingByEducation %>% filter(`parental level of education` == "master's degree") %>% pull(avg_writing_score)


``` 

```{r}


 education <- c('some high school', 'high school', 'some college', "associate's degree", "bachelor's degree", "master's degree")

 newSP |>

    filter(variable == 'parental level of education') |>

    ggplot(aes(y=mean_score, x = factor(category, education), fill=category))+

    geom_boxplot(color = "black")+

    labs(y = "Average Score",x='', title = "Level of Parent's Education")+

    theme_minimal()+

    theme(

    axis.title.x = element_blank(),

    axis.text.x = element_blank(),

    axis.ticks.x = element_blank()

    )

```

## Results
As we see there is a general trend upwards in all scores comapring with the education level of the parents

notably those whose parents who did finish highscool did not score as well as those who did not finsih highschool.

Children of parents with Masters are noticalbly ahead of those with bachelors

Little change between finishing highschool and a bachelors degree

# Effects of School Lunch Programs on scores

```{r}

newSP |>

    filter(variable == 'lunch') |>

    ggplot(aes(y=mean_score, x = category, fill=category))+

    geom_boxplot(color = "black")+

    labs(y = "Average Score",x='', title = "Type of Lunch")+

    theme_minimal()+

    theme(

    axis.title.x = element_blank(),

    axis.text.x = element_blank(),

    axis.ticks.x = element_blank()

    )

```


## Results

Students who were not on a school lunch assistance program scored higher than those who were

# Conclusion

- While there are marginal differences between many factors

  - Test preperation had a large effect

  - Ethnicity had a large effect

  - Lunch status had a large effect

- The trend here shows that the greatest factor of how the student would do is likely their parents income

- while we dont have access to that data directly, we have strong reasoning

  - Test preperation courses are expensive

  - Being on a lunch program is generally an indicator of a low income household

  - There is a similar distribution in Ethicity and test score and ethnicity and wealth
