---
title: "What Factors Determine a Students Success?"
---


# What we did

## Chose Data:

-   we really wanted to find a data set that would apply to us and could give us results that we would care about

-   We were all interested to see how a students environment effected their success in school

-   Results could be used to paint a better picture for where external funding should go for students falling behind. 

## Clean Data

-   This data set was relatively small, only 1000 entries, so there really was not that much garbage to clean out

-   all exam scores ranged from 0-100, and every student had a full attribute list.

## Goals

-   The goal of our project was to use the data to determine which attributes of the students were most important

-   There are 5 attributes in this data set 

    -   Gender

    -   Ethnicity

    -   Education Level of the students parents

    -   Whether they were enrolled in some sort of school lunch program

    -   If they took any preparation courses for the exam

-   We needed to compare all of these to decide which one was the most important factor in each of the 3 subjects and overall. 

## Process Data

-   First we needed some baseline averages for each category so we have a reference to work with

-   We used R to go through each column and find an average score for each section, and a average total score

-   We then modeled each attribute to show its frequency in the data so we had a better idea as to what the population looked like

-   

```{r}
#| echo: true
#| eval: true

library(ggplot2)

students <- readr::read_csv("../datas/StudentsPerformance.csv")
readingMean <- mean(students$'reading score', na.rm = TRUE)
writingMean <- mean(students$'writing score', na.rm = TRUE)
mathMean <- mean(students$'math score', na.rm = TRUE)
```
```{r}
#| echo: false
print(students)
```

#### Reading Mean:
```{r}
#| echo: false
print(readingMean)
```

#### Writing Mean: 
```{r}
#| echo: false
print(writingMean)
```

#### Math Mean:
```{r}
#| echo: false
print(mathMean)

```

### Now that we have all of our means, we can start seeing where the variations lie within attributes

```{r}
library(dplyr)
print(ReadingByPrep<-students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))
)
avg_reading_completed <- ReadingByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_reading_score)
avg_reading_none <- ReadingByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_reading_score)
```



```{r}
library(dplyr)
MathByPrep <- students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

avg_math_completed <- MathByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_math_score)
avg_math_none <- MathByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_math_score)
```


```{r}
WritingByPrep <- students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

avg_writing_completed <- WritingByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_writing_score)
avg_writing_none <- WritingByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_writing_score)
```

## Here are the results:
-   There was a `r avg_reading_completed - avg_reading_none`% increase for reading scores when prep courses were taken

-   There was a `r avg_writing_completed - avg_writing_none`% increase for writing scores when prep courses were taken

-   There was a `r avg_math_completed - avg_math_none`% increase for math scores when prep courses were taken

# Effect of Gender on Scores

```{r}

ReadingByGender<-students %>%
  group_by(`gender`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByGender<-students %>%
  group_by(`gender`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByGender<-students %>%
    group_by(`gender`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

print(MathByGender)
print(ReadingByGender)
print(WritingByGender)

avg_writing_male <- WritingByGender %>% filter(`gender` == "male") %>% pull(avg_writing_score)
avg_writing_female <- WritingByGender %>% filter(`gender` == "female") %>% pull(avg_writing_score)

avg_reading_male <- ReadingByGender %>% filter(`gender` == "male") %>% pull(avg_reading_score)
avg_reading_female <- ReadingByGender %>% filter(`gender` == "female") %>% pull(avg_reading_score)

avg_math_male <- MathByGender %>% filter(`gender` == "male") %>% pull(avg_math_score)
avg_math_female <- MathByGender %>% filter(`gender` == "female") %>% pull(avg_math_score)


print(avg_math_male)
print(avg_math_female)
print(avg_reading_male)
print(avg_reading_female)
print(avg_writing_male)
print(avg_writing_female)


