---
title: "What Factors Determine a Students Success?"
---


# What we did

## Chose Data:

-   we really wanted to find a data set that would apply to us and could give us results that we would care about

-   We were all interested to see how a students environment effected their success in school

-   Results could be used to paint a better picture for where external funding should go for students falling behind. 

## Clean Data

-   This data set was relatively small, only 1000 entries, so there really was not that much garbage to clean out

-   all exam scores ranged from 0-100, and every student had a full attribute list.

## Goals

-   The goal of our project was to use the data to determine which attributes of the students were most important

-   There are 5 attributes in this data set 

    -   Gender

    -   Ethnicity

    -   Education Level of the students parents

    -   Whether they were enrolled in some sort of school lunch program

    -   If they took any preparation courses for the exam

-   We needed to compare all of these to decide which one was the most important factor in each of the 3 subjects and overall. 

## Process Data

-   First we needed some baseline averages for each category so we have a reference to work with

-   We used R to go through each column and find an average score for each section, and a average total score

-   We then modeled each attribute to show its frequency in the data so we had a better idea as to what the population looked like

-   

```{r}
#| echo: true
#| eval: true

library(ggplot2)

students <- readr::read_csv("./datas/StudentsPerformance.csv")
readingMean <- mean(students$'reading score', na.rm = TRUE)
writingMean <- mean(students$'writing score', na.rm = TRUE)
mathMean <- mean(students$'math score', na.rm = TRUE)
```
```{r}
#| echo: false
print(students)
```

#### Reading Mean:
```{r}
#| echo: false
print(readingMean)
```

#### Writing Mean: 
```{r}
#| echo: false
print(writingMean)
```

#### Math Mean:
```{r}
#| echo: false
print(mathMean)

```

## Now that we have all of our means, we can start seeing where the variations lie within attributes



# Effects of Prep Courses on Scores

```{r}
library(dplyr)
print(ReadingByPrep<-students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))
)
avg_reading_completed <- ReadingByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_reading_score)
avg_reading_none <- ReadingByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_reading_score)
```



```{r}
library(dplyr)
MathByPrep <- students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

avg_math_completed <- MathByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_math_score)
avg_math_none <- MathByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_math_score)
```


```{r}
WritingByPrep <- students %>%
  group_by(`test preparation course`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

avg_writing_completed <- WritingByPrep %>% filter(`test preparation course` == "completed") %>% pull(avg_writing_score)
avg_writing_none <- WritingByPrep %>% filter(`test preparation course` == "none") %>% pull(avg_writing_score)

prep_levels <- c("Completed", "None")

avg_reading_scores <- c(avg_reading_completed, avg_reading_none)
avg_writing_scores <- c(avg_writing_completed, avg_writing_none)
avg_math_scores <- c(avg_math_completed, avg_math_none)

df <- tibble(
  Prep_Course = factor(rep(prep_levels, 3), levels = prep_levels), 
  Subject = rep(c("Reading", "Writing", "Math"), each = length(prep_levels)),  
  Score = c(avg_reading_scores, avg_writing_scores, avg_math_scores)  
)

ggplot(df, aes(x = Prep_Course, y = Score, fill = Subject)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Test Preparation Course",
       x = "Test Preparation Course",
       y = "Average Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```





-   There was a `r avg_reading_completed - avg_reading_none`% increase for reading scores when prep courses were taken

-   There was a `r avg_writing_completed - avg_writing_none`% increase for writing scores when prep courses were taken

-   There was a `r avg_math_completed - avg_math_none`% increase for math scores when prep courses were taken

# Effect of Gender on Scores

```{r}

ReadingByGender<-students %>%
  group_by(`gender`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByGender<-students %>%
  group_by(`gender`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByGender<-students %>%
    group_by(`gender`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

print(MathByGender)
print(ReadingByGender)
print(WritingByGender)

avg_writing_male <- WritingByGender %>% filter(`gender` == "male") %>% pull(avg_writing_score)
avg_writing_female <- WritingByGender %>% filter(`gender` == "female") %>% pull(avg_writing_score)

avg_reading_male <- ReadingByGender %>% filter(`gender` == "male") %>% pull(avg_reading_score)
avg_reading_female <- ReadingByGender %>% filter(`gender` == "female") %>% pull(avg_reading_score)

avg_math_male <- MathByGender %>% filter(`gender` == "male") %>% pull(avg_math_score)
avg_math_female <- MathByGender %>% filter(`gender` == "female") %>% pull(avg_math_score)


genders <- c("Male", "Female")

avg_reading_scores <- c(avg_reading_male, avg_reading_female)
avg_writing_scores <- c(avg_writing_male, avg_writing_female)
avg_math_scores <- c(avg_math_male, avg_math_female)


df <- tibble(
  Gender = factor(rep(genders, 3), levels = genders),  # Set factor levels explicitly
  Subject = rep(c("Reading", "Writing", "Math"), each = length(genders)),  # Repeat subjects
  Score = c(avg_reading_scores, avg_writing_scores, avg_math_scores)  # Combine all scores
)

ggplot(df, aes(x = Gender, y = Score, fill = Subject)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Gender",
       x = "Gender",
       y = "Average Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))





```

## Results

The effects of gender on test scores are very interesting

- In Mathematics, the males pull ahead by `r avg_math_male - avg_math_female`%

- In Reading and writing, the females win by a large gap

  - Reading: `r avg_reading_female - avg_reading_male`% win for females

  - Writing: `r avg_writing_female - avg_writing_male`% win for females


# Effect of Ethnicity on Scores

#### For this data set, there are 5 Ethnicities listed, they are Listed ethnicities A, B, C, D, E


- To determine whether certain ethnicities accel in certain subjects, we need to compare an 

- ethnicity's average score for each subject and the global score for that subject

```{r}

library(dplyr)

ReadingByEthnicity<-students %>%
  group_by(`race/ethnicity`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByEthnicity<-students %>%
  group_by(`race/ethnicity`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByEthnicity<-students %>%
    group_by(`race/ethnicity`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))

avgMathEthnicityA<-MathByEthnicity %>% filter(`race/ethnicity` == "group A") %>% pull(avg_math_score)
avgMathEthnicityB<-MathByEthnicity %>% filter(`race/ethnicity` == "group B") %>% pull(avg_math_score)
avgMathEthnicityC<-MathByEthnicity %>% filter(`race/ethnicity` == "group C") %>% pull(avg_math_score)
avgMathEthnicityD<-MathByEthnicity %>% filter(`race/ethnicity` == "group D") %>% pull(avg_math_score)
avgMathEthnicityE<-MathByEthnicity %>% filter(`race/ethnicity` == "group E") %>% pull(avg_math_score)

avgReadingEthnicityA<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group A") %>% pull(avg_reading_score)
avgReadingEthnicityB<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group B") %>% pull(avg_reading_score)
avgReadingEthnicityC<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group C") %>% pull(avg_reading_score)
avgReadingEthnicityD<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group D") %>% pull(avg_reading_score)
avgReadingEthnicityE<-ReadingByEthnicity %>% filter(`race/ethnicity` == "group E") %>% pull(avg_reading_score)

avgWritingEthnicityA<-WritingByEthnicity %>% filter(`race/ethnicity` == "group A") %>% pull(avg_writing_score)
avgWritingEthnicityB<-WritingByEthnicity %>% filter(`race/ethnicity` == "group B") %>% pull(avg_writing_score)
avgWritingEthnicityC<-WritingByEthnicity %>% filter(`race/ethnicity` == "group C") %>% pull(avg_writing_score)
avgWritingEthnicityD<-WritingByEthnicity %>% filter(`race/ethnicity` == "group D") %>% pull(avg_writing_score)
avgWritingEthnicityE<-WritingByEthnicity %>% filter(`race/ethnicity` == "group E") %>% pull(avg_writing_score)

ethnicities <- c("A", "B", "C", "D","E")
avg_reading_scores <- c(avgReadingEthnicityA, avgReadingEthnicityB, avgReadingEthnicityC, avgReadingEthnicityD, avgReadingEthnicityE)
avg_writing_scores <- c(avgWritingEthnicityA, avgWritingEthnicityB, avgWritingEthnicityC, avgWritingEthnicityD, avgWritingEthnicityE)
avg_math_scores <- c(avgMathEthnicityA, avgMathEthnicityB, avgMathEthnicityC, avgMathEthnicityD, avgMathEthnicityE)

df <- tibble(
  Ethnicity = factor(rep(ethnicities, 3), levels = ethnicities),  # Set factor levels explicitly
  Subject = rep(c("Reading", "Writing", "Math"), each = length(ethnicities)),  # Repeat subjects
  Score = c(avg_reading_scores, avg_writing_scores, avg_math_scores)  # Combine all scores
)

ggplot(df, aes(x = Ethnicity, y = Score, fill = Subject)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Ethnicity",
       x = "Ethnicity",
       y = "Average Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


print(ethnicities)
print(avg_math_scores)
print(avg_reading_scores)
print(avg_writing_scores)

print(mathMean)
print(readingMean)
print(writingMean)

```

- someone add something more about this im done for now

# Effects of Parents Education Level on Scores
#### there are in this data set: 
#### some highschool, highschool, some college, associates, bachelors, masters

 we will now compare the scores based off of the parents education level.

```{r}
ReadingByEducation<-students %>%
  group_by(`parental level of education`) %>%
  summarise(avg_reading_score = mean(`reading score`, na.rm = TRUE))

WritingByEducation<-students %>%
  group_by(`parental level of education`) %>%
  summarise(avg_writing_score = mean(`writing score`, na.rm = TRUE))

MathByEducation<-students %>%
    group_by(`parental level of education`) %>%
    summarise(avg_math_score = mean(`math score`, na.rm = TRUE))



avgMathsomeHighschool<-MathByEducation %>% filter(`parental level of education` == "some high school") %>% pull(avg_math_score)
avgReadingsomeHighschool<-ReadingByEducation %>% filter(`parental level of education` == "some high school") %>% pull(avg_reading_score)
avgWritingsomeHighschool<-WritingByEducation %>% filter(`parental level of education` == "some high school") %>% pull(avg_writing_score)

avgMathsHighschool<-MathByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_math_score)
avgReadingHighschool<-ReadingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_reading_score)
avgWritingHighschool<-WritingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_writing_score)

avgMathHighschool<-MathByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_math_score)
avgReadingHighschool<-ReadingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_reading_score)
avgWritingHighschool<-WritingByEducation %>% filter(`parental level of education` == "high school") %>% pull(avg_writing_score)

avgMathSomeCollege<-MathByEducation %>% filter(`parental level of education` == "some college") %>% pull(avg_math_score)
avgReadingSomeCollege<-ReadingByEducation %>% filter(`parental level of education` == "some college") %>% pull(avg_reading_score)
avgWritingSomeCollege<-WritingByEducation %>% filter(`parental level of education` == "some college") %>% pull(avg_writing_score)

avgMathAssociates<-MathByEducation %>% filter(`parental level of education` == "associate's degree") %>% pull(avg_math_score)
avgReadingAssociates<-ReadingByEducation %>% filter(`parental level of education` == "associate's degree") %>% pull(avg_reading_score)
avgWritingAssociates<-WritingByEducation %>% filter(`parental level of education` == "associate's degree") %>% pull(avg_writing_score)

avgMathBachelors<-MathByEducation %>% filter(`parental level of education` == "bachelor's degree") %>% pull(avg_math_score)
avgReadingBachelors<-ReadingByEducation %>% filter(`parental level of education` == "bachelor's degree") %>% pull(avg_reading_score)
avgWritingBachelors<-WritingByEducation %>% filter(`parental level of education` == "bachelor's degree") %>% pull(avg_writing_score)

avgMathMasters<-MathByEducation %>% filter(`parental level of education` == "master's degree") %>% pull(avg_math_score)
avgReadingMasters<-ReadingByEducation %>% filter(`parental level of education` == "master's degree") %>% pull(avg_reading_score)
avgWritingMasters<-WritingByEducation %>% filter(`parental level of education` == "master's degree") %>% pull(avg_writing_score)


``` 
Whole Lot of brute force in those calculations but we have our data, now to the graphs...

```{r}
readingByEducation <- c(avgReadingsomeHighschool, avgReadingHighschool, avgReadingSomeCollege, avgReadingAssociates, avgReadingBachelors, avgReadingMasters)
writingByEducation <- c(avgWritingsomeHighschool, avgWritingHighschool, avgWritingSomeCollege, avgWritingAssociates, avgWritingBachelors, avgWritingMasters)
mathByEducation <- c(avgMathsomeHighschool, avgMathHighschool, avgMathSomeCollege, avgMathAssociates, avgMathBachelors, avgMathMasters)
education_levels <- c("Some High School", "High School", "Some College", "Associates", "Bachelors", "Masters")

df <- tibble(
  Education = factor(rep(education_levels, 3), levels = education_levels),  # Set factor levels explicitly
  Subject = rep(c("Reading", "Writing", "Math"), each = length(education_levels)),  # Repeat subjects
  Score = c(readingByEducation, writingByEducation, mathByEducation)  # Combine all scores
)

ggplot(df, aes(x = Education, y = Score, fill = Subject)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Scores by Education Level",
       x = "Education Level",
       y = "Average Score") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

As we see there is a general trend upwards in all scores comapring with the education level of the parents

notably those whose parents who did finish highscool did not score as well as those who did not finsih highschool.




