---
title: "What Factors Determine a Students Success?"
---


# What we did

## Chose Data:

-   we really wanted to find a data set that would apply to us and could give us results that we would care about

-   We were all interested to see how a students environment effected their success in school

-   Results could be used to paint a better picture for where external funding should go for students falling behind. 

## Clean Data

-   This data set was relatively small, only 1000 entries, so there really was not that much garbage to clean out

-   all exam scores ranged from 0-100, and every student had a full attribute list.

## Goals

-   The goal of our project was to use the data to determine which attributes of the students were most important

-   There are 5 attributes in this data set 

    -   Gender

    -   Ethnicity

    -   Education Level of the students parents

    -   Whether they were enrolled in some sort of school lunch program

    -   If they took any preparation courses for the exam

-   We needed to compare all of these to decide which one was the most important factor in each of the 3 subjects and overall. 

## Process Data

-   First we needed some baseline averages for each category so we have a reference to work with

-   We used R to go through each column and find an average score for each section, and a average total score

-   We then modeled each attribute to show its frequency in the data so we had a better idea as to what the population looked like

-   

```{r}
#| echo: true
#| eval: true

library(ggplot2)

students <- readr::read_csv("./datas/StudentsPerformance.csv")
readingMean <- mean(students$'reading score', na.rm = TRUE)
writingMean <- mean(students$'writing score', na.rm = TRUE)
mathMean <- mean(students$'math score', na.rm = TRUE)
```
```{r}
#| echo: false
print(students)
```

## Mean Scores

- Mean math score `r mathMean` 

- Mean Reading score `r readingMean`

- Mean writing score `r writingMean`

## Now that we have all of our means, we can start seeing where the variations lie within attributes



# Effects of Prep Courses on Scores
